Explicit Jacobian matrix formulas for entropy stable summation-by-parts schemes

### 2020-04-03 docs/ESDG_jacobian/main.tex
updating docs

Both space-time and implicit time discretizations require the solution of a system of nonlinear equations.  This can be done using Newton's method, which involves the Jacobian matrix of the nonlinear equations.  While it is possible to compute the solution to the nonlinear system without explicitly computing the Jacobian matrix using Jacobian-free Newton-Krylov methods \cite{knoll2004jacobian, birken2019subcell}, the Jacobian matrix is also useful for preconditioning \cite{persson2006efficient, persson2008newton}.  In this work, we present explicit formulas for Jacobian matrices of systems resulting from entropy stable formulations.  We also show that computing the Jacobian matrix is not significantly more expensive than evaluating the residual of the nonlinear system.  Finally, we apply the new Jacobian formulas to both explicit two-derivative and implicit time-stepping schemes.

@@

For clarity of presentation, we consider first a scalar nonlinear conservation law in one spatial dimension
\eqlab{
\pd{u}{t} + \pd{f(u)}{x} = 0. \label{eq:ncl}
}
We assume without loss of generality periodic boundary conditions, which will simplify the presentation of the main results.  Extensions to non-periodic boundaries are treated in Section~\ref{eq:npbc}.  The generalization to systems of nonlinear conservation laws and higher dimensions is also postponed until Section~\ref{sec:systems}.

@@

The aforementioned proofs require partial derivatives of flux functions $f_S(u_L,u_R)$ with respect to at least one argument.  While this can be done by hand for simple fluxes, it rapidly becomes cumbersome for complex or piecewise-defined flux functions such as the logarithmic mean \cite{ismail2009affordable, winters2019entropy}.  This can be avoided by using Automatic Differentiation (AD) \cite{griewank2008evaluating}.  AD is distinct from both symbolic differentiation and finite difference approximations in that it does not return an explicit expression for derivatives, but is still able to evaluate derivatives accurately up to machine precision.  Given a computational function, AD returns a second black-box function which evaluates the derivative.

In this work, we utilize the Julia implementation of forward-mode automatic differentiation provided by \verb+ForwardDiff.jl+ \cite{RevelsLubinPapamarkou2016}.  The procedure is remarkably simple: given some flux function \verb+f(x,y)+, \verb+ForwardDiff.jl+ returns the derivative with respect to either $x$ or $y$ as another function.  For example, defining the function $\LRl{\pd{f}{y}}_{x,y}$ is a one-line operation:
\begin{verbbox}
dfdy(x,y) = ForwardDiff.derivative(y->f(x,y),y)
\end{verbbox}
\begin{figure}[!h]
\centering
\theverbbox
\end{figure}
\\
This simple API is made possible by the flexible Julia type system  \cite{bezanson2017julia}.

In principle, AD can be directly applied to $\bm{f}(\bm{u})$ to compute the Jacobian matrix.  However, because AD scales with the number of inputs and outputs, the cost of applying AD directly to $\bm{f}(\bm{u})$ increases as the discretization resolution increases.  In contrast, using the approach in this paper, the entries of the Jacobian matrix can be computed for roughly the same cost as a single evaluation of $\bm{f}(\bm{u})$.  Moreover, explicit Jacobian formulas enable implementations to take advantage of sparsity in $\bm{Q}$ when assembling the Jacobian matrix.

@@

These matrices can be used to construct entropy conservative high order discretizations.  Let $J = h/2$ is the Jacobian of the mapping from the reference element $[-1,1]$ to a physical interval of size $h$ and let $\bm{F}^k_{ij} = f_S(u_{i,k},u_{j,k})$ denote the matrix of flux interactions between different nodes on the element $D^k$.  A local formulation on the element $D^k$ is given by
\eqlab{
J_k {\bm{M}} \td{\bm{u}_k}{t} + 2\LRp{{\bm{Q}} \circ \bm{F}^k}\bm{1} + \bm{B}\LRp{\bm{f}^*-f(\bm{u}_k)}   = \bm{0}, \qquad \bm{f}^* = \bmat{
f_S(u_{1,k}^+,u_{1,k})\\
0\\
\vdots\\
f_S(u_{N_p,k}^+,u_{N_p,k})
},
\label{eq:sbp1D}
}
where $u_{1,k}^+, u_{N_p,k}^+$ denote the exterior values of $u_{1,k}, u_{N_p,k}$ on neighboring elements.  Assuming that the elements are ordered from left to right in ascending order, for interior element indices $1 < k < K$, these are given by
\[
u_{1,k}^+ = u_{N_p,k-1}, \qquad u_{N_p,k}^+ = u_{1,k+1}.
\]
In other words, the first node on $D^k$ is connected to the last node on the previous element, and the last node on $D^k$ is connected to the first node on the next element.

@@

The generalization to higher dimensional domains and curved geometric mappings is straightforward, but notationally much more complicated.  The construction of skew matrices $\bm{Q}_{\Omega}$ follows from approaches detailed in \cite{crean2018entropy, chan2017discretely, chan2018discretely, chan2018efficient, chan2019skew, hicken2020entropy}.  We omit it for conciseness in this work.

@@

Let $\bm{u}_{\Omega}$ denote a vector of vectors 
\[
\bm{u}_{\Omega} = \bmat{\bm{u}_1\\
\bm{u}_2\\
\vdots\\
\bm{u}_n}, \qquad \bm{u}_i = \bmat{
\bm{u}_{i,1}\\
\bm{u}_{i,2}\\
\vdots\\
\bm{u}_{i,K}
}, \qquad
\bm{u}_{i,k} = \bmat{
\bm{u}_{i,k,1}\\
\bm{u}_{i,k,2}\\
\vdots\\
\bm{u}_{i,k,N_p}
}
\]
for $\ell = 1,\ldots, n$, $k = 1,\ldots, K$ and $j = 1,\ldots, N_p$.  Here, $\bm{u}_{\ell,k,j}$ denotes the $j$th degree of freedom for the $\ell$th component of the solution on the $k$th element.  
%We redefine these matrices as the $n \times n$ block matrices 
%\eq{
%&\bm{M}_{\Omega} \longrightarrow \bm{I}_n\otimes \bm{M}_{\Omega} = \bmat{\bm{M}_{\Omega} &  & \\
%& \ddots & \\
% &  & \bm{M}_{\Omega}},
%\\
%&\bm{Q}_{\Omega} \longrightarrow \bm{e}_n\bm{e}_n^T\otimes \bm{Q}_{\Omega} = 
%\bmat{\bm{Q}_{\Omega} & \ldots & \bm{Q}_{\Omega}\\
%\vdots & \ddots & \vdots\\
%\bm{Q}_{\Omega} & \ldots & \bm{Q}_{\Omega}},
%}
%where $\bm{e}_n\in \mathbb{R}^n$ is the vector of all ones, and $\bm{e}_n\bm{e}_n^T$ is the $n\times n$ matrix of all ones.  
Let $(k_1, j_1)$ and $(k_2, j_2)$ be tuples which correspond to row and column indices of a matrix.  
We now define the flux matrix $\bm{F}$ as the $n\times n$ block matrix containing flux interactions between different nodes
\eqlab{
\bm{F} = \bmat{
\bm{F}_1 &&\\
& \ddots &\\
&& \bm{F}_n
}, \qquad \LRp{\bm{F}_\ell}_{(k_1, j_1),(k_2,j_2)} = \LRp{\bm{f}_{S}\LRp{\bm{u}_{:, k_1,j_1},\bm{u}_{:, k_2,j_2}}}_{\ell}.
\label{eq:fluxsys}
}
where $\bm{u}_{:,k,j}$ denotes the vector containing all solution components at the $k$th element and $j$th node, and each block $\bm{F}_\ell$ for $\ell = 1,\ldots, n$ corresponds to the $\ell$th component of the vector-valued flux evaluated at solution states $\bm{u}_{:,k_1,j_1},\bm{u}_{:,k_2,j_2}$.

Let $\bm{M}_{\Omega}, \bm{Q}_{\Omega}$ denote the global mass and differentiation matrices in (\ref{eq:sbpmat}).  Then, an entropy conservative scheme is given by
\[
\LRp{\bm{I}_n\otimes \bm{M}_{\Omega}}\td{\bm{u}_{\Omega}}{t} + 2\LRp{ \LRp{\bm{e}_n\bm{e}_n^T\otimes \bm{Q}_{\Omega} } \circ \bm{F}}\bm{1} = \bm{0}.
\]
where $\bm{e}_n\in \mathbb{R}^n$ is the vector of all ones, and $\bm{e}_n\bm{e}_n^T$ is the $n\times n$ matrix of all ones.

We now construct explicit Jacobian matrix formulas for systems of nonlinear conservation laws.  The right hand side function $\bm{f}(\bm{u})$ for systems can be rewritten as
\[
\bm{f}(\bm{u}) = 2\LRp{ \bm{Q}_{\Omega} \circ \bm{F}}\bm{1} 
= 2\bmat{(\bm{Q}_{\Omega}\circ \bm{F}_1) \\ 
\vdots\\
(\bm{Q}_{\Omega}\circ \bm{F}_n)
}\bm{1}
\]
Then, the Jacobian matrix is 
\eqlab{
\pd{\bm{f}}{\bm{u}} &= 2\bmat{
\bm{Q}_{\Omega}\circ \bm{F}_{1,\bm{u}_1} & \ldots & \bm{Q}_{\Omega}\circ \bm{F}_{1,\bm{u}_n}\\ 
\vdots & \ddots & \vdots\\
\bm{Q}_{\Omega}\circ \bm{F}_{n,\bm{u}_1} & \ldots & \bm{Q}_{\Omega}\circ \bm{F}_{n,\bm{u}_n}
} \label{eq:fluxjac}\\
&= \LRp{\bm{e}_n\bm{e}_n^T\otimes \bm{Q}_{\Omega}} \circ 
\bmat{
\bm{F}_{1,\bm{u}_1} & \ldots &\bm{F}_{1,\bm{u}_{n}}\\
\vdots & \ddots & \vdots \\
\bm{F}_{n,\bm{u}_1} & \ldots &\bm{F}_{n,\bm{u}_n}\nonumber
}.
}
Here, the flux matrices $\bm{F}_{i,\bm{u}_j}$ are evaluated via
\[
\LRp{\bm{F}_{i,\bm{u}_j}}_{j_1k_1,j_2k_2} = \LRl{\pd{\LRp{\bm{f}_S}_i}{\bm{u}_j}}_{\bm{u}_{:,k_1,j_1},\bm{u}_{:,k_2,j_2}},
\]
where $\pd{\LRp{\bm{f}_S}_i}{\bm{u}_{j}}$ denotes the derivative of the $i$th component of the flux $\bm{f}_S\LRp{\bm{u}_L,\bm{u}_R}$ with respect to the $j$th solution component of $\bm{u}_R$.  Thus, each block $\bm{F}_{i,\bm{u}_j}$ corresponds to an entry of the Jacobian of $\bm{f}_S(\bm{u}_L,\bm{u}_R)$).

\begin{remark}
We note that an alternative ordering of degrees of freedom \cite{crean2018entropy} yields a Jacobian matrix with a more compact bandwidth.   A smaller matrix bandwidth can also be achieved by bandwidth-minimizing algorithms such as the reverse Cuthill-McKee algorithm \cite{cuthill1969reducing}.
%Let $\bm{u}_{\Omega}$ now denote a vector of vectors 
%\[
%\bm{u}_{\Omega} = \bmat{\bm{u}_1\\
%\bm{u}_2\\
%\vdots\\
%\bm{u}_K}, \qquad \bm{u}_k = \bmat{
%\bm{u}_{k,1}\\
%\bm{u}_{k,2}\\
%\vdots\\
%\bm{u}_{k,N_p}
%}, \qquad
%\bm{u}_{k,j} = \bmat{
%\bm{u}_{k,j,1}\\
%\bm{u}_{k,j,2}\\
%\vdots\\
%\bm{u}_{k,j,n}
%}
%\]
%for $k = 1,\ldots, K$ and $j = 1,\ldots, N_p$.  Here, $\bm{u}_{k,j,i}$ denotes the $j$th degree of freedom for the $i$th component of the solution on the $k$th element.   Define global mass and differentiation matrices $\bm{M}_{\Omega}, \bm{Q}_{\Omega}$ as in (\ref{eq:sbpmat}), with the local blocks $\bm{M}, \bm{Q},\bm{B}_L, \bm{B}_R$ replaced by 
%\eq{
%\bm{M} &\longrightarrow \bm{M} \otimes \bm{I}_n, \qquad 
%\bm{Q} \longrightarrow \bm{Q} \otimes \bm{I}_n,\\
%\bm{B}_L &\longrightarrow \bm{B}_L \otimes \bm{I}_n, \qquad 
%\bm{B}_R \longrightarrow \bm{B}_R \otimes \bm{I}_n,
%}
%where $\bm{I}_n$ is the $n\times n$ identity matrix.  Note that $\bm{Q}_{\Omega}$ as defined in (\ref{eq:sbpmat}) remains skew-symmetric under this new definition.  We also redefine the flux matrix to account for the vector-valued nature of the flux function $\bm{f}_S$ 
%\eqlab{
%\bm{F} = \bmat{
%\bm{F}_{11} & \ldots & \bm{F}_{1K}\\
%\vdots & \ddots & \vdots\\
%\bm{F}_{K1} & \ldots & \bm{F}_{KK}
%}, \qquad \LRp{\bm{F}_{k_1,k_2}}_{ij} = \diag{\bm{f}_S(\bm{u}_{k_1,i}, \bm{u}_{k_2,j})}.\label{eq:fluxsystem}
%}
%Then, an entropy conservative formulation is given by
%\[
%\bm{M}_{\Omega}\td{\bm{u}_{\Omega}}{t} + 2\LRp{ \bm{Q}_{\Omega} \circ \bm{F}}\bm{1} = \bm{0}.
%\]
%Explicit Jacobians can also be constructed for this ordering.  We omit it for conciseness.
\end{remark}

@@

\section{Generalizations to different numerical settings}

In this section, we discuss how to generalize the explicit construction of Jacobians to different numerical settings, including non-collocation schemes and non-periodic boundary conditions.

\subsection{Non-collocated schemes: hybridized SBP operators, entropy projection, and over-integration}

@@

\subsubsection{Entropy conservative schemes and hybridized SBP operators}

@@

We can also use $\bm{Q}_h$ to construct a globally skew-symmetric differentiation matrix (see also \cite{chan2018efficient}).  Define the matrix $\bm{S} = \bm{Q}-\bm{Q}^T$, and define $\bm{Q}_{\Omega}$ as the global block matrix
\[
\bm{Q}_{\Omega} = \frac{1}{2}\LRs{\begin{array}{cc|cc|cc|cc}
\bm{S} &  \bm{E}^T\bm{B} & &&& && \\
 - \bm{B}\bm{E} &  & & \bm{B}_R && && -\bm{B}_L\\ [.2em]
 \hline\\[-1.0em]
&& \bm{S} &  \bm{E}^T\bm{B}&&&& \\
& -\bm{B}_L &  - \bm{B}\bm{E} &&& \bm{B}_R&&\\
 \hline\\[-1.0em]
&&&& \ddots & \ddots&  \\
&&&   -\bm{B}_L & \ddots&\ddots&   \ddots&\\
 \hline\\[-1.0em]
&&&&&\ddots& \bm{S} &  \bm{E}^T\bm{B} \\
\bm{B}_R &&& &&-\bm{B}_L &  - \bm{B}\bm{E} &\\
\end{array}},
\]
We also construct global interpolation, projection, and extrapolation matrices
\eq{
\bm{V} &\longrightarrow \bm{I}_K \otimes \bm{V}, \qquad \bm{P} \longrightarrow \bm{I}_K \otimes \bm{P} \\ 
\bm{E} &\longrightarrow \bm{I}_K \otimes \bm{E}, \qquad \bm{V}_h \longrightarrow \bm{I}_K \otimes \bm{V}_h
}
Finally, we assume that the global solution $\bm{u}(\bm{x},t) \in \mathbb{R}^n$ is vector-valued, and order the solution coefficients as in Section~\ref{sec:systems}.

>>>>

Both space-time and implicit time discretizations require the solution of a system of nonlinear equations.  This can be done using Newton's method, which involves the Jacobian matrix of the nonlinear equations.  While it is possible to compute the solution to the nonlinear system without explicitly computing the Jacobian matrix using Jacobian-free Newton-Krylov methods \cite{knoll2004jacobian, birken2019subcell}, the Jacobian matrix is commonly used to construct preconditioners \cite{persson2008newton}.  In this work, we present explicit formulas for Jacobian matrices of systems resulting from entropy stable formulations.  We also show that computing the Jacobian matrix is not significantly more expensive than evaluating the residual of the nonlinear system.  Finally, we apply the new Jacobian formulas to both explicit two-derivative and implicit time-stepping schemes.

@@

For clarity of presentation, we consider first a scalar nonlinear conservation law in one spatial dimension
\eqlab{
\pd{u}{t} + \pd{f(u)}{x} = 0. \label{eq:ncl}
}
We assume without loss of generality periodic boundary conditions, which will simplify the presentation of the main results.  
%Extensions to non-periodic boundaries are treated in Section~\ref{eq:npbc}.  
The generalization to systems of nonlinear conservation laws and higher dimensions is postponed until Section~\ref{sec:systems}.

@@

The aforementioned proofs require partial derivatives of flux functions $f_S(u_L,u_R)$ with respect to at least one argument.  While this can be done by hand for simple fluxes, it rapidly becomes cumbersome for complex or piecewise-defined flux functions such as the logarithmic mean \cite{ismail2009affordable, winters2019entropy}.  This can be avoided by using Automatic Differentiation (AD) \cite{griewank2008evaluating}.  AD is distinct from both symbolic differentiation and finite difference approximations in that it does not return an explicit expression for derivatives, but is still able to evaluate derivatives accurately up to machine precision.  Given a computational function, AD provides a black-box function which evaluates the derivative.

In this work, we utilize the Julia implementation of forward-mode automatic differentiation provided by \verb+ForwardDiff.jl+ \cite{RevelsLubinPapamarkou2016}.  The procedure is remarkably simple: given some flux function \verb+f(x,y)+, \verb+ForwardDiff.jl+ returns the derivative with respect to either $x$ or $y$ as another function.  For example, defining the function $\LRl{\pd{f}{y}}_{x,y}$ is a one-line operation:
\begin{verbbox}
dfdy(x,y) = ForwardDiff.derivative(y->f(x,y),y)
\end{verbbox}
\begin{figure}[!h]
\centering
\theverbbox
\end{figure}
\\
The computation of Jacobians for vector-valued flux functions can be performed similarly using \verb+ForwardDiff.jacobian+.  This simple API is made possible by the flexible Julia type system \cite{bezanson2017julia}.

In principle, AD can be directly applied to $\bm{f}(\bm{u})$ to compute the Jacobian matrix.  However, because AD scales with the number of inputs and outputs, the cost of applying AD directly to $\bm{f}(\bm{u})$ increases as the discretization resolution increases.  In contrast, using the approach in this paper, AD is applied only to the flux function, which has a small number of inputs and outputs which are fixed independently of the discretization resolution.  As a result, the cost of evaluating derivatives of the flux function is roughly the same cost as evaluating the flux function, and entries of the Jacobian matrix can be computed for roughly the same cost as a single evaluation of the nonlinear flux $\bm{f}(\bm{u})$.  Moreover, explicit Jacobian formulas enable implementations to take advantage of sparsity in $\bm{Q}$ when assembling the Jacobian matrix.

@@

These matrices can be used to construct entropy conservative high order discretizations.  Let $J = h/2$ is the Jacobian of the mapping from the reference element $[-1,1]$ to a physical interval of size $h$ and let $\bm{F}^k_{ij} = f_S(u_{i,k},u_{j,k})$ denote the matrix of flux interactions between different nodes on the element $D^k$.  A local formulation on the element $D^k$ is given by
\begin{gather}
J_k {\bm{M}} \td{\bm{u}_k}{t} + 2\LRp{{\bm{Q}} \circ \bm{F}^k}\bm{1} + \bm{B}\LRp{\bm{f}^*-f(\bm{u}_k)}  = \bm{0}, \nonumber\\
\bm{f}^* = \bmat{
f_S(u_{1,k}^+,u_{1,k})\\
0\\
\vdots\\
0\\
f_S(u_{N_p,k}^+,u_{N_p,k})
}.
\label{eq:sbp1D}
\end{gather}
where $u_{1,k}^+, u_{N_p,k}^+$ denote the exterior values of $u_{1,k}, u_{N_p,k}$ on neighboring elements.  Assuming that the elements are ordered from left to right in ascending order, for interior element indices $1 < k < K$, these are given by
\[
u_{1,k}^+ = u_{N_p,k-1}, \qquad u_{N_p,k}^+ = u_{1,k+1}.
\]
In other words, the first node on $D^k$ is connected to the last node on the previous element, and the last node on $D^k$ is connected to the first node on the next element.

@@

The generalization to higher dimensional domains and curved geometric mappings is straightforward, but notationally much more complicated.  The construction of skew matrices $\bm{Q}_{\Omega}$ follows from approaches detailed in \cite{crean2018entropy, chan2017discretely, chan2018discretely, chan2018efficient, chan2019skew, hicken2020entropy}.  We omit it for conciseness in this work, but note that the efficient assembly of DG matrices is discussed in detail elsewhere \cite{persson2008newton, hillewaert2011sharp, kempf2018automatic}.

@@

Let $\bm{u}_{\Omega}$ denote a vector of vectors 
\[
\bm{u}_{\Omega} = \bmat{\bm{u}_1\\
\bm{u}_2\\
\vdots\\
\bm{u}_n}, \qquad \bm{u}_i = \bmat{
\bm{u}_{i,1}\\
\bm{u}_{i,2}\\
\vdots\\
\bm{u}_{i,K}
}, \qquad
\bm{u}_{i,k} = \bmat{
\bm{u}_{i,k,1}\\
\bm{u}_{i,k,2}\\
\vdots\\
\bm{u}_{i,k,N_p}
}
\]
for $\ell = 1,\ldots, n$, $k = 1,\ldots, K$ and $j = 1,\ldots, N_p$.  Here, $\bm{u}_{\ell,k,j}$ denotes the $j$th degree of freedom for the $\ell$th component of the solution on the $k$th element.  
%We redefine these matrices as the $n \times n$ block matrices 
%\eq{
%&\bm{M}_{\Omega} \longrightarrow \bm{I}_n\otimes \bm{M}_{\Omega} = \bmat{\bm{M}_{\Omega} &  & \\
%& \ddots & \\
% &  & \bm{M}_{\Omega}},
%\\
%&\bm{Q}_{\Omega} \longrightarrow \bm{e}_n\bm{e}_n^T\otimes \bm{Q}_{\Omega} = 
%\bmat{\bm{Q}_{\Omega} & \ldots & \bm{Q}_{\Omega}\\
%\vdots & \ddots & \vdots\\
%\bm{Q}_{\Omega} & \ldots & \bm{Q}_{\Omega}},
%}
%where $\bm{e}_n\in \mathbb{R}^n$ is the vector of all ones, and $\bm{e}_n\bm{e}_n^T$ is the $n\times n$ matrix of all ones.  
Let $(k_1, j_1)$ and $(k_2, j_2)$ be multi-indices which correspond to row and columns indices of a matrix, respectively.  
We now define the flux matrix $\bm{F}$ as the $n\times n$ block diagonal matrix containing flux interactions between different nodes
\eqlab{
\bm{F} = \bmat{
\bm{F}_1 &&\\
& \ddots &\\
&& \bm{F}_n
}, \qquad \LRp{\bm{F}_\ell}_{(k_1, j_1),(k_2,j_2)} = \LRp{\bm{f}_{S}\LRp{\bm{u}_{:, k_1,j_1},\bm{u}_{:, k_2,j_2}}}_{\ell}.
\label{eq:fluxsys}
}
where $\bm{u}_{:,k,j}$ denotes the vector containing all solution components at the $k$th element and $j$th node, and each entry of the block $\bm{F}_\ell$ for $\ell = 1,\ldots, n$ corresponds to the $\ell$th component of the vector-valued flux evaluated at solution states $\bm{u}_{:,k_1,j_1},\bm{u}_{:,k_2,j_2}$.

Let $\bm{M}_{\Omega}, \bm{Q}_{\Omega}$ denote the global mass and differentiation matrices in (\ref{eq:sbpmat}).  Then, an entropy conservative scheme is given by
\[
\LRp{\bm{I}_n\otimes \bm{M}_{\Omega}}\td{\bm{u}_{\Omega}}{t} + 2\LRp{ \LRp{\bm{I}_n\otimes \bm{Q}_{\Omega} } \circ \bm{F}}\bm{1} = \bm{0}.
\]
where $\bm{e}_n\in \mathbb{R}^n$ is the vector of all ones, and $\bm{e}_n\bm{e}_n^T$ is the $n\times n$ matrix of all ones.

We now construct explicit Jacobian matrix formulas for systems of nonlinear conservation laws.  The right hand side function $\bm{f}(\bm{u})$ for systems can be rewritten as
\[
\bm{f}(\bm{u}) = 2\LRp{ \LRp{\bm{I}_n\otimes\bm{Q}_{\Omega}} \circ \bm{F}}\bm{1} 
= 2\bmat{(\bm{Q}_{\Omega}\circ \bm{F}_1) \\ 
\vdots\\
(\bm{Q}_{\Omega}\circ \bm{F}_n)
}\bm{1}
\]
Then, the Jacobian matrix is 
\eqlab{
\pd{\bm{f}}{\bm{u}} &=
\bmat{
\bm{\partial F}_{1,\bm{u}_1} & \ldots & \bm{\partial F}_{1,\bm{u}_n}\\
\vdots & \ddots & \vdots\\
\bm{\partial F}_{n,\bm{u}_1} & \ldots & \bm{\partial F}_{n,\bm{u}_n}
}
% 2\bmat{
%\bm{Q}_{\Omega}\circ \bm{F}_{1,\bm{u}_1} & \ldots & \bm{Q}_{\Omega}\circ \bm{F}_{1,\bm{u}_n}\\ 
%\vdots & \ddots & \vdots\\
%\bm{Q}_{\Omega}\circ \bm{F}_{n,\bm{u}_1} & \ldots & \bm{Q}_{\Omega}\circ \bm{F}_{n,\bm{u}_n}
%} + \text{\note{add diagonal terms }}\label{eq:fluxjac}\\
%&= \LRp{\bm{e}_n\bm{e}_n^T\otimes \bm{Q}_{\Omega}} \circ 
%\bmat{
%\bm{F}_{1,\bm{u}_1} & \ldots &\bm{F}_{1,\bm{u}_{n}}\\
%\vdots & \ddots & \vdots \\
%\bm{F}_{n,\bm{u}_1} & \ldots &\bm{F}_{n,\bm{u}_n}\nonumber
%}.
}
where the Jacobian block $\bm{\partial F}_{i,\bm{u}_j}$ is evaluated as in Lemma~\ref{lemma:explicitJ}
\[
\bm{\partial F}_{i,\bm{u}_j} = 2 \LRp{\bm{Q}_{\Omega} \circ \bm{F}_{i,\bm{u}_j}} \pm \diag{\bm{1}^T\LRp{2\bm{Q}_{\Omega}\circ \bm{F}_{i,\bm{u}_j}}}.
\]
for $\bm{Q}_{\Omega} = \pm \bm{Q}_{\Omega}^T$.  
Here, the flux matrix $\bm{F}_{i,\bm{u}_j}$ is evaluted via
\[
\LRp{\bm{F}_{i,\bm{u}_j}}_{j_1k_1,j_2k_2} = \LRl{\pd{\LRp{\bm{f}_S}_i}{\bm{u}_{R,j}}}_{\bm{u}_{:,k_1,j_1},\bm{u}_{:,k_2,j_2}},
\]
where $\pd{\LRp{\bm{f}_S}_i}{\bm{u}_{R,j}}$ denotes the derivative of the $i$th component of the flux $\bm{f}_S\LRp{\bm{u}_L,\bm{u}_R}$ with respect to the $j$th solution component of $\bm{u}_R$.  Thus, each entry of the block $\bm{F}_{i,\bm{u}_j}$ corresponds to an entry of the Jacobian (with respect to $\bm{u}_R$) of $\bm{f}_S(\bm{u}_L,\bm{u}_R)$ and an entry of the global differentiation matrix $\bm{Q}_{\Omega}$.

\begin{remark}
An alternative ordering of degrees of freedom \cite{crean2018entropy} yields a Jacobian matrix with a more compact bandwidth.   A smaller matrix bandwidth can also be achieved by precomputing bandwidth-minimizing row and column permutations using the reverse Cuthill-McKee algorithm \cite{cuthill1969reducing}.  
%Let $\bm{u}_{\Omega}$ now denote a vector of vectors 
%\[
%\bm{u}_{\Omega} = \bmat{\bm{u}_1\\
%\bm{u}_2\\
%\vdots\\
%\bm{u}_K}, \qquad \bm{u}_k = \bmat{
%\bm{u}_{k,1}\\
%\bm{u}_{k,2}\\
%\vdots\\
%\bm{u}_{k,N_p}
%}, \qquad
%\bm{u}_{k,j} = \bmat{
%\bm{u}_{k,j,1}\\
%\bm{u}_{k,j,2}\\
%\vdots\\
%\bm{u}_{k,j,n}
%}
%\]
%for $k = 1,\ldots, K$ and $j = 1,\ldots, N_p$.  Here, $\bm{u}_{k,j,i}$ denotes the $j$th degree of freedom for the $i$th component of the solution on the $k$th element.   Define global mass and differentiation matrices $\bm{M}_{\Omega}, \bm{Q}_{\Omega}$ as in (\ref{eq:sbpmat}), with the local blocks $\bm{M}, \bm{Q},\bm{B}_L, \bm{B}_R$ replaced by 
%\eq{
%\bm{M} &\longrightarrow \bm{M} \otimes \bm{I}_n, \qquad 
%\bm{Q} \longrightarrow \bm{Q} \otimes \bm{I}_n,\\
%\bm{B}_L &\longrightarrow \bm{B}_L \otimes \bm{I}_n, \qquad 
%\bm{B}_R \longrightarrow \bm{B}_R \otimes \bm{I}_n,
%}
%where $\bm{I}_n$ is the $n\times n$ identity matrix.  Note that $\bm{Q}_{\Omega}$ as defined in (\ref{eq:sbpmat}) remains skew-symmetric under this new definition.  We also redefine the flux matrix to account for the vector-valued nature of the flux function $\bm{f}_S$ 
%\eqlab{
%\bm{F} = \bmat{
%\bm{F}_{11} & \ldots & \bm{F}_{1K}\\
%\vdots & \ddots & \vdots\\
%\bm{F}_{K1} & \ldots & \bm{F}_{KK}
%}, \qquad \LRp{\bm{F}_{k_1,k_2}}_{ij} = \diag{\bm{f}_S(\bm{u}_{k_1,i}, \bm{u}_{k_2,j})}.\label{eq:fluxsystem}
%}
%Then, an entropy conservative formulation is given by
%\[
%\bm{M}_{\Omega}\td{\bm{u}_{\Omega}}{t} + 2\LRp{ \bm{Q}_{\Omega} \circ \bm{F}}\bm{1} = \bm{0}.
%\]
%Explicit Jacobians can also be constructed for this ordering.  We omit it for conciseness.
\end{remark}

@@

%\section{Generalizations to different numerical settings}

%In this section, we discuss how to generalize the explicit construction of Jacobians to different numerical settings, including non-collocation schemes and non-periodic boundary conditions.

\section{Non-collocated schemes: hybridized SBP operators, entropy projection, over-integration}

@@

\subsection{Entropy conservative schemes and hybridized SBP operators}

@@

We can also use $\bm{Q}_h$ to construct a globally skew-symmetric differentiation matrix (see also \cite{chan2018efficient}).  Define the matrix $\bm{S} = \bm{Q}-\bm{Q}^T$, and define $\bm{Q}_{\Omega}$ as the global block matrix
\[
\bm{Q}_{\Omega} = \frac{1}{2}\LRs{\begin{array}{cc|cc|cc|cc}
\bm{S} &  \bm{E}^T\bm{B} & &&& && \\
 - \bm{B}\bm{E} &  & & \bm{B}_R && && -\bm{B}_L\\ [.2em]
 \hline\\[-1.0em]
&& \bm{S} &  \bm{E}^T\bm{B}&&&& \\
& -\bm{B}_L &  - \bm{B}\bm{E} &&& \bm{B}_R&&\\
 \hline\\[-1.0em]
&&&& \ddots & \ddots&  \\
&&&   -\bm{B}_L & \ddots&\ddots&   \ddots&\\
 \hline\\[-1.0em]
&&&&&\ddots& \bm{S} &  \bm{E}^T\bm{B} \\
\bm{B}_R &&& &&-\bm{B}_L &  - \bm{B}\bm{E} &\\
\end{array}},
\]
We also abuse notation and redefine $\bm{V},\bm{E},\bm{P},\bm{V}_h$ as \emph{global} interpolation, projection, and extrapolation matrices
\eq{
\bm{V} &\longrightarrow \bm{I}_K \otimes \bm{V}, \qquad \bm{P} \longrightarrow \bm{I}_K \otimes \bm{P} \\ 
\bm{E} &\longrightarrow \bm{I}_K \otimes \bm{E}, \qquad \bm{V}_h \longrightarrow \bm{I}_K \otimes \bm{V}_h
}
Finally, we assume that the global solution $\bm{u}(\bm{x},t) \in \mathbb{R}^n$ is vector-valued, and order the solution coefficients as in Section~\ref{sec:systems}.

It was shown in \cite{parsani2016entropy, chan2017discretely} that, when either the mass matrix is non-diagonal or the nodal set does not contain appropriate boundary points, it is necessary to perform an entropy projection (or extrapolation \cite{chenreview}) step to ensure discrete entropy stability.  Let $\bm{v}(\bm{u})$ denote the entropy variables as a function of the conservative variables, and let $\bm{u}(\bm{v})$ denote the inverse mapping.  We define the entropy projected variables $\tilde{\bm{u}}_{\Omega}$ as
\eqlab{
\tilde{\bm{u}}_{\Omega} = \bm{u}\LRp{\bm{V}_h\bm{P}\bm{v}\LRp{\bm{V}\hat{\bm{u}}_{\Omega}}}.
\label{eq:entropyprojection}
}
Let $\bm{F}$ again denote the block-diagonal flux matrix in (\ref{eq:fluxsys}).  We evaluate each flux block $\bm{F}_{\ell}$  using the entropy projected variables
\eqlab{
\LRp{\bm{F}_\ell}_{(k_1, j_1),(k_2,j_2)} = \LRp{\bm{f}_{S}\LRp{\tilde{\bm{u}}_{:, k_1,j_1},\tilde{\bm{u}}_{:, k_2,j_2}}}_{\ell}.
\label{eq:fluxell}
}
Then, an entropy conservative method is given by 
\[
\LRp{\bm{I}_n\otimes \bm{M}_{\Omega}}\td{\bm{u}_{\Omega}}{t} + 2\LRp{\bm{I}_n\otimes \bm{V}_h}^T\LRp{ \LRp{\bm{e}_n\bm{e}_n^T\otimes \bm{Q}_{\Omega} } \circ \bm{F}}\bm{1} = \bm{0}.
%\bm{M}_{\Omega}\td{\bm{u}_{\Omega}}{t} + 2\bm{V}^T\LRp{ \bm{Q}_{\Omega} \circ \bm{F}}\bm{1} = \bm{0}.  
\]
where $\bm{I}_n$ is the $n\times n$ identity matrix.

\subsection{Jacobian matrices for hybridized SBP schemes}

We redefine the nonlinear term as
\[
\bm{f}(\hat{\bm{u}}) = 2\LRp{\bm{I}_n\otimes \bm{V}_h}^T\LRp{ \LRp{\bm{e}_n\bm{e}_n^T\otimes \bm{Q}_{\Omega}} \circ \bm{F}}\bm{1}.
\]
where the flux matrix $\bm{F}$ is computed using the entropy projected conservative variables (\ref{eq:entropyprojection}) via (\ref{eq:fluxell}).
Let $\pd{\bm{u}}{\bm{v}}$ and $\pd{\bm{v}}{\bm{u}}$ denote Jacobians of the conservative variables with respect to the entropy variables and vice versa.  These have been explicitly derived for several equations (for example, the Jacobians for the compressible Navier-Stokes equations are given in \cite{hughes1986new}).
